---
title: 'TASK 3: Santa Barbara urchins'
author: "Jaxon Stuhr"
date: "10/25/2021"
output: html_document
---
## Task Overview

This code wrangles and analyzes data on Santa Barbara urchins. Two species, red urchins and purple urchins, are compared at 5 sites around Santa Barbara, CA, on the basis of observed size and prevalence. 

**Data citation:**

Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 


```{r setup, include=TRUE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(lubridate)
```

```{r}
# read in urchins data 
urchins = read.csv(here("data", "urchins.csv"))
```

## Section 1: Size comparisons, purple and red urchins

### Data Visualization
A comparison of the size distribution for purple and red urchins.

```{r}
# plot using urchins data, x var = size
ggplot(data = urchins, aes(x = SIZE, fill = COMMON_NAME)) + 
  # histogram w 12 bins
  geom_histogram(bins = 12) + 
  # facet wrap based on common name
  facet_wrap(~ COMMON_NAME) +
  # manually control colors
  scale_fill_manual(values=c("Purple", "Dark Red")) +
  # add x, y labels and title
  labs(x = "Urchin Diameter [cm]", 
       y = "Count", 
       title = "Size Distribution for Purple and Red Sea Urchins") +
  # change theme
  theme_linedraw() +
  # center title
  theme(plot.title = element_text(hjust = 0.5)) +
  # remove legend
  guides(fill = FALSE) 

```

Fig 1: Urchin diameter distribution [cm] for red and purple urchins in Santa Barbara, CA

### Descriptive statistics table
Building a summary table including the mean, median, standard deviation, and sample size and sorting by species (red or purple urchin).

```{r}
urchin_stats = urchins %>% 
  # group by common name of urchins (red or purple)
  group_by(COMMON_NAME) %>% 
  summarize(
    # add summary stats mean, median, sd, sample size
    mean_size = round(mean(SIZE),1),
    median_size = median(SIZE),
    sd_size = round(sd(SIZE),1),
    sample_size = n()
  )

urchin_stats %>%
  # build table w kbl extra, rename columns
  kbl(caption = "Table 1: Santa Barbara Urchins Summary Statistics", col.names = c("Common Name", "Mean Diameter [cm]", "Median Diameter [cm]", "Standard Deviation [cm]", "Sample Size")) %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

### Section 1 Analysis

```{r}
# calculate difference
mean_size_dif = (urchin_stats$mean_size[2] - urchin_stats$mean_size[1])
# calculate percent difference between sample means
mean_size_percent_dif = 
  round(
    200 * 
      (urchin_stats$mean_size[2] - urchin_stats$mean_size[1]) /
      (urchin_stats$mean_size[2] + urchin_stats$mean_size[1]), 
    1)
#min and max red urchin size
reds = urchins %>% 
  filter(COMMON_NAME == "Red Urchin") 
min_red = min(reds$SIZE)
max_red = max(reds$SIZE)
#min and max purple urchin size
purples = urchins %>% 
  filter(COMMON_NAME == "Purple Urchin") 
min_purple = min(purples$SIZE)
max_purple = max(purples$SIZE)

```

The red urchins were, on average `r mean_size_dif` cm larger and  `r mean_size_percent_dif` % larger in diameter than the purple urchins. The red urchins also had a much larger spread, with diameters ranging from `r min_red` to `r max_red` cm, whereas the purple urchins diamters ranged from `r min_purple` to `r max_purple` cm.

## Section 2: Exploring purple urchin counts by site and year

### Pre-processing the purple urchins data

```{r}
# start w urchins
purple_urchins = urchins %>% 
  # only purple ones
  filter(COMMON_NAME == "Purple Urchin") %>% 
  # use libridate on DATE
  mutate(DATE = mdy(DATE)) %>% 
  # add yr column
  mutate(YEAR = year(DATE)) %>% 
  # add column w full site names
  mutate(site_full = case_when(
    SITE == "NAPL" ~ "Naples Reef",
    SITE == "AQUE" ~ "Arroyo Quemado Reef",
    SITE == "CARP" ~ "Carpinteria Reef",
    SITE == "MOHK" ~ "Mohawk Reef",
    SITE == "IVEE" ~ "Isla Vista Reef"
  ))
```

### Building a table to store the total number of purple urchins observed by site and year

```{r, include=FALSE}
# start w purples
purple_urchins_by_site_year = purple_urchins %>% 
  #group by site and yr
  group_by(site_full, YEAR) %>% 
  # count em
  count()
```

### Plotting counts of purple urchins recorded at each site by year

```{r}
# plot w x =  yr, y = count
ggplot(purple_urchins_by_site_year, aes(x = YEAR, y = n, fill = site_full)) + 
  # column graph
  geom_col() +
  # break up by site
  facet_wrap(~ site_full) + 
  # add labels
  labs(x = "Year", y = "Number of Observed Urchins", title = "Observed Urchins by Site and Year") +
  # change theme
  theme_linedraw() + 
  # center title
  theme(plot.title = element_text(hjust = 0.5)) +
  # remove legend
  guides(fill = FALSE) 
```

Fig 2: Observed Urchin counts for purple urchins at 5 sites around Santa Barbara, CA for years 2008 - 2020

### Purple Urchins Analysis

The purple urchin counts decreased at all sites other than the Isla Vista Reef site, appearing to portray a population decline starting around 2012. We must also, however, take into account the sampling mechanism and insure that sampling procedures were constant across observed years. Even if this is the case, it is also possible that the urchins may be present in higher densities in portions of the reefs that were not sampled. 







